Fitting metabolic models to dissolved oxygen data: the Estuarine BAyesian Single-station Estimation method

### 2023-05-02 manu-draft.qmd
added text about synthetic plus noise methods and results

As a final evaluation of EBASE to return known metabolic parameters, a new time series was created that added estimated effects of instrument and tidal noise to the synthetic time series.  This assessment provided information on the influence of common noise signals in dissolved oxygen time series on the estimation process for EBASE.  An assumption of the analysis is that EBASE will return similar parameters for a noisy time series as for those in the time series used above that did not include noise.  The instrument and tidal noise time series were first estimated by using a weighted regression analysis [@Beck15] on an actual observed dissolved oxygen time series for Apalachicola Bay using data from 2021.

@@

@fig-apacmpfig and `r run_reference(id = "tbl-apacmptab")` provide a comparison of the metabolic estimates from the EBASE and Odum methods. Correlations for all metabolic estimates, regardless of the input time series, were positive and significantly correlated, as expected.  However, correlations were generally stronger for the metabolic results based on the detided dissolved oxygen time series.  The only exception for the correlation estimates was gas exchange (D), which had a slightly weaker correlation using the detided results. @fit-apacmpfig shows that the Odum results for gas exchange were more variable for the same values returned as no or small gas exchange for the EBASE method based on the detided dissolved oxygen. The RMSE values of each regression comparing the metabolic estimates between methods were generally lower in the detided results, excluding D which was higher with the detided dissolved oxygen relative to the observed.

The comparisons in @fig-apacmpfig and `r run_reference(id = "tbl-apacmptab")` suggest that tidal effects on continuous monitoring data can severely violate assumptions required for methods of estimating ecosystem metabolism [as described in detail in @Beck15]. These effects are easily identified by stronger correlations and lower RMSE using the detided dissolved oxygen.  Further, effects of tidal advection can be seen with estimates from the Odum method as negative values for P and R, i.e., "anomalous" values. The effects of tidal advection on estimates from EBASE were more subtle as anomalous values cannot be returned based on constraints of the priors (i.e., none can be negative), although the range of values for the metabolic estimates were much larger with observed dissolved oxygen. Overall, the results suggest that detiding the dissolved oxygen time series prior to estimating metabolic rates reduces tidal noise and produces more stable estimates with smaller ranges, regardless of the method used.

@@

-   Primary modifications (Production is linearly related to PAR (aPAR), not saturation point Respiration is constant (not temp dependent), Uses a gas exchange commonly used in oceanic systems (not littoral - streams/lakes as in Grace))
-   Describe how EBASE deals with gaps and flexibility of the model to accommodate diff optimization periods
-   b is sketchy, how to improve? Either fix or develop JAGS model further to include additional data, e.g. tidal height.
-   detiding
-   other main conclusions (from RGN comments on draft?)

@@

![Comparison of EBASE results to a synthetic time series created with known metabolic parameters. Results are compared for unique combinations of prior distributions and optimization period. The means ($\mu$) and standard deviations ($\sigma$) are evaluated as low (L) or high (H) values (see text for details) and the optimization period is one day, seven days, or thirty days. The metabolic results for dissolved oxygen (DO), gross primary production (P), ecosystem respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values. NSE values from zero to one (green) indicate EBASE has recovered the synthetic parameters with values closer to 1 being a better approximation, whereas negative values indicate the mean of the synthetic data is a better predictor than EBASE.  The color scale for NSE is also exponentially transformed to minimize the influence of large negative values on the visual display. The $b$ parameter is a fixed constant and was not evaluated.  Rankings of each combination as the median NSE value across all parameters in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/priorcomp.png){#fig-priorcomp}

>>>>

As a final evaluation of EBASE to return known metabolic parameters, a new time series was created that added estimated effects of instrument and tidal noise to the synthetic time series. The objective of the analysis is to develop an expectation of how metabolic estimates may change using in situ data with noise that may substantially mask a biological signal in dissolved oxygen measurements. The instrument and tidal noise were estimated using a weighted regression analysis [@Beck15; @Beck21] on actual observed dissolved oxygen time series for Apalachicola Bay using data from 2021.  This approach models dissolved oxygen as a function of time, salinity, and season using a moving window approach with weights of the independent variables appropriate for each time period within the windows.  The results of this analysis return a predicted and detided dissolved oxygen time series.  The instrument noise was estimated as the difference between the observed and predicted time series and the tidal noise was estimated as the difference between the predicted and detided time series.  Both the instrument and tidal noise were added to the same synthetic time series above (@fig-synapanos).  EBASE was then used with unconstrained priors (large $\sigma$) on the synthetic time series with noise, then compared with results from the original synthetic time series.

@fig-synapanoscmp shows the comparison of the EBASE results for the synthetic time series with instrument and tidal noise to those from the original synthetic time series. Results for both time series produced similar dissolved oxygen estimates (top-left panel, @fig-synapanoscmp), although notable differences were observed in the remaining parameters.  In particular, gross primary production (P) and ecosystem respiration (R) were generally higher and more variable for results from the noisy time series, with differences in the former attributed to higher estimates in the $a$ parameter.  As for the previous results, the $b$ parameter was estimated with low precision for both input time series, although results were centered around the true value of 0.251 $\buni$.  Gas exchange (D) was similar between the input time series, suggesting lack of precision in the $b$ parameter regardless of noise had a minimal influence.

@@

@fig-apacmpfig and `r run_reference(id = "tbl-apacmptab")` provide a comparison of the metabolic estimates from the EBASE and Odum methods. Correlations for all metabolic estimates, regardless of the input time series, were positive and significantly correlated, as expected.  However, correlations were generally stronger for the metabolic results based on the detided dissolved oxygen time series.  The only exception for the correlation estimates was gas exchange (D), which had a slightly weaker correlation using the detided results. @fig-apacmpfig shows that the Odum results for gas exchange were more variable for the same values returned as no or small gas exchange for the EBASE method based on the detided dissolved oxygen. The RMSE values of each regression comparing the metabolic estimates between methods were generally lower in the detided results, excluding D which was higher with the detided dissolved oxygen relative to the observed.

The comparisons in @fig-apacmpfig and `r run_reference(id = "tbl-apacmptab")` suggest that tidal effects on continuous monitoring data can severely violate assumptions required for methods of estimating ecosystem metabolism [as described in detail in @Beck15]. These effects are easily identified by stronger correlations and lower RMSE using the detided dissolved oxygen.  Further, effects of tidal advection can be seen with estimates from the Odum method as negative values for P and R, i.e., "anomalous" values. The effects of tidal advection on estimates from EBASE were more subtle as anomalous values cannot be returned based on constraints of the priors (i.e., none can be negative), although the range of values for the metabolic estimates were much larger with observed dissolved oxygen. A similar result (larger estimates for P and R) was observed using the synthetic time series with instrument and tidal noise (@fig-synapanoscmp). Overall, the results suggest that detiding the dissolved oxygen time series prior to estimating metabolic rates reduces tidal noise and produces more stable estimates with smaller ranges, regardless of the method used.

@@

-   Primary modifications (Production is linearly related to PAR (aPAR), not saturation point Respiration is constant (not temp dependent), Uses a gas exchange commonly used in oceanic systems (not littoral - streams/lakes as in Grace))
-   Describe how EBASE deals with gaps and flexibility of the model to accommodate diff optimization periods
-   b is sketchy, how to improve? Either fix or develop JAGS model further to include additional data, e.g. tidal height.
-   detiding - noise may produce higher and more variable metabolic rates
-   other main conclusions (from RGN comments on draft?)

@@

![Comparison of EBASE results to a synthetic time series created with known metabolic parameters. Results are compared for unique combinations of prior distributions and optimization period. The means ($\mu$) and standard deviations ($\sigma$) are evaluated as low (L) or high (H) values (see text for details) and the optimization period for EBASE is one day, seven days, or thirty days. The metabolic results for dissolved oxygen (DO), gross primary production (P), ecosystem respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values. NSE values from zero to one (green) indicate EBASE has recovered the synthetic parameters with values closer to 1 being a better approximation, whereas negative values indicate poor recovery.  The color scale for NSE is also exponentially transformed to minimize the influence of large negative values on the visual display. The $b$ parameter is a fixed constant and was not evaluated.  Rankings of each combination as the median NSE value across all parameters in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/priorcomp.png){#fig-priorcomp}
